
### Exercise 2.3 {#sec-prob_post_weibull_lung}

*This exercise continues the analysis of the lung cancer data in Exercise 2.2*

Assume that the survival time $X$ of the lung cancer patients in Exercise 2.2 are independent Weibull distributed 
$$
X_1,\ldots,X_n \vert \lambda, k \overset{\mathrm{iid}}{\sim} \mathrm{Weibull}(\lambda,k).
$$
The value of $k$ determines how the failure rate changes with time:

- $k=1$ gives a failure (death) rate that is constant over time and corresponds to the special case of a exponential distribution $\mathrm{Expon}(\theta=1/\lambda)$ used in Exercise 2.2. Note that (following Wikipedia) the exponential distribution is parameterized with a rate (inverse scale) parameter $\theta$, while the Weibull is parameterized with a scale parameter $\lambda= 1/\theta$ 🤷
- $k<1$ gives a decreasing failure rate over time
- $k>1$ gives an increasing failure rate over time.

(a) Plot the posterior distribution of $\lambda$ conditional on $k=1$, $k=3/2$ and $k=2$. For all $k$, use the prior $\lambda \sim \mathrm{Gamma}(\alpha,\beta)$ with $\alpha=3$ and $\beta=1/50$ (which a similar prior for $\theta=1/\lambda$ as in Exercise 2.2). *Hint*: the posterior distribution for $k\neq 1$ is intractable, so use numerical evaluation of the posterior over a grid of $\lambda$-values.
(b) Plot the `time` variable as a histogram and overlay the fitted model for the three different $k$-values; use the posterior mode for $\theta$ in each model when plotting the fitted model density.<br>


::: {.callout-note icon="false" collapse="true"}
## Solution Exercise 2.3a

Similar to Exercise 2.2b, the likelihood can be computed with separate treatment of the uncensored and censored observations: 
$$
\begin{align}
p(x_1,\ldots,x_n \vert \lambda, k) & = \prod_{i=1}^n p(x_i \vert \lambda, k) \\
& = \prod_{u \in \mathcal{U}} p(x_u \vert \lambda, k) \prod_{c \in \mathcal{C}} \Big(1 - F(x_c \vert \lambda, k)\Big) 
\end{align}
$$
where $p(x \vert \lambda, k)$ is the pdf of a Weibull variable
$$
p(x \vert \lambda, k) = \frac{k}{\lambda}\Big( \frac{x}{\lambda} \Big)^{k-1}e^{-(x/\lambda)^k}\quad\text{ for }x>0
$$
which is implemented in R as `dweibull`. The cdf of the Weibull distribution is of rather simple form
$$
F(x \vert \lambda, k) = 1 - e^{-(x/\lambda)^k}
$$
and is implemented in R as `pweibull`.

The code below plots the prior and posterior distribution for $\lambda$ for the three different $k$-values. We could have inserted the mathematical expressions for the pdf and cdf and simplified the final likelihood expression; we will instead use the `dweibull` and `pweibull` functions without simplifications since it gives a more general template that can be used for any distribution, not just the Weibull model. For numerical stability we usually compute the posterior distribution on the log scale
$$
\log p(\lambda^{(j)} \vert x_1,\ldots,x_n) \propto \log p(x_1,\ldots,x_n \vert \lambda_j) + \log p(\lambda_j)
$$
for a grid of equally spaced $\lambda$-values: $\lambda^{(1)}\ldots,\lambda^{(J)}$. The $\propto$ sign now means that there is a missing *additive* constant $\log p(x_1,\ldots,x_n)$ which does not depend on the unknown parameter $\lambda$. When we have computed $\log p(\lambda \vert x_1,\ldots,x_n)$ over a grid of $\lambda$ values we compute the posterior on the original scale by
$$
p(\lambda^{(j)} \vert x_1,\ldots,x_n) \propto \exp\Big( \log p(x_1,\ldots,x_n \vert \lambda_j) + \log p(\lambda_j) \Big)
$$
and then divide all numbers with the normalizing constant to make sure that the posterior integrates to one. This is done numerically by approximating the integral by a Riemann rectangle sum 
$$
p(\lambda^{(j)} \vert x_1,\ldots,x_n) = 
\frac{\exp\Big( \log p(x_1,\ldots,x_n \vert \lambda^{(j)}) + \log p(\lambda^{(j)}) \Big)}
{\sum_{h=1}^J \exp\Big( \log p(x_1,\ldots,x_n \vert \lambda^{(h)}) + \log p(\lambda^{(h)}) \Big) \Delta}
$$
where $\Delta$ is the spacing between the grid points of $\lambda$-values: $\lambda^{(1)}, \ldots, \lambda^{(J)}$.

```{r}
#| output: false
library(tidyverse) # loads data manipulation and visualization packages
library(survival) # loads the lung cancer data as `lung`
colors = c("#6C8EBF", "#c0a34d", "#780000","#007878","#B5C6DF","#EADAAA","#AE6666")
```

Set up prior hyperparameters
```{r}
alpha_prior <- 3     # shape parameter
beta_prior <- 1/300  # rate parameter
```

Set up function that computes the likelihood for any $\lambda$ value:
```{r}
# Make a function that computes the likelihood
weibull_loglike <- function(lambda, x, censored, k){
  loglik_uncensored = sum(dweibull(x[-censored], shape = k, scale = lambda, 
                                   log = TRUE))
  loglik_censored = sum(pweibull(x[censored], shape = k, scale = lambda, 
                                 lower.tail = FALSE, log.p = TRUE))
  return(loglik_uncensored + loglik_censored)
}
```

Set up a function that computes the posterior density over a grid of $\lambda$:
```{r}
weibull_posterior <- function(lambdaGrid, x, censored, k, alpha_prior, beta_prior){
  Delta = lambdaGrid[2] - lambdaGrid[1] # Grid step size
  logPrior <- dgamma(lambdaGrid, shape = alpha_prior, rate = beta_prior, log = TRUE)
  logLike <- sapply(lambdaGrid, weibull_loglike, x, censored, k)
  logPost <- logLike + logPrior
  logPost <- logPost - max(logPost) # subtract constant to avoid overflow
  post <- exp(logPost)/(sum(exp(logPost))*Delta) # original scale and normalize
  logLike <- logLike - max(logLike)
  likeNorm <- exp(logLike)/(sum(exp(logLike))*Delta) # normalized likelihood
  return(list(post = post, prior = exp(logPrior), likeNorm = likeNorm))
}
```

```{r}
# Plot the prior and posterior densities

lambdaGrid <- seq(1, 1000, length.out = 1000)
# Compute to get the prior
postRes <- weibull_posterior(lambdaGrid, lung$time, lung$status == 1, k = 1, 
                             alpha_prior, beta_prior)
df <- data.frame(
  lambdaGrid = lambdaGrid, 
  prior = postRes$prior
)

# Compute for all selected k values
postModes = c()
for (k in c(1, 3/2, 2)){
  postRes <- weibull_posterior(lambdaGrid, lung$time, lung$status == 1, k, alpha_prior, beta_prior)
  df[str_glue("posterior k={k}")] <- postRes$post
  postModes = c(postModes, lambdaGrid[which.max(postRes$post)])
}

df_long <- df %>% pivot_longer(-lambdaGrid, names_to = "density_type", values_to = "density")

# Plot using ggplot2
ggplot(df_long) +
  aes(x = lambdaGrid, y = density, color = density_type) +
  geom_line() +
  scale_colour_manual(
    breaks = c("prior", "posterior k=1", "posterior k=1.5", "posterior k=2"), 
    values = c(colors[2], colors[1], colors[3], colors[4])) +
  labs(title = "Exercise 2.3", x = expression(lambda), y = "Density", color = "") + 
  theme_minimal()
```
:::


::: {.callout-note icon="false" collapse="true"}
## Solution Exercise 2.3b
The fit of the three Weibull models are plotted below. The best fit seems to be for $k=3/2$, which is still not very good. In a later exercise you will be asked to freely estimate both $\lambda$ and $k$.
```{r}
ggplot(lung, aes(time)) +
  geom_histogram(aes(y = after_stat(density), fill = "Data"), bins = 30) +
  stat_function(fun = dweibull, args = list(shape = 1, scale = postModes[1]), lwd = 1, 
                aes(color = "Weibull fit k = 1/2"),
  ) +
  stat_function(fun = dweibull, args = list(shape = 3/2, scale = postModes[2]), lwd = 1, 
                aes(color = "Weibull fit k = 1"),
  ) +
  stat_function(fun = dweibull, args = list(shape = 2, scale = postModes[3]), lwd = 1, 
                aes(color = "Weibull fit k = 2"),
  ) +
  labs(title = "Weibull model fits", x = "days", y = "Density") + 
  scale_fill_manual("", values = colors[6]) +
  scale_color_manual("", values = c(colors[1], colors[3], colors[4])) +
  theme_minimal()

```

:::
