
### Exercise 2.3
The dataset `lung` in the R package `survival` contains data on 228 patients with advanced lung cancer. We will here analyze the survival time in days for the patients which is recorded by the variable `time`. The variable `status` is a binary variable with `status = 1` if the survival time of the patient is censored (patient still alive at the end of the study) and `status = 2` if the survival time was uncensored (patient dead before the end of the study). 

Assume that the survival time $X$ of the patients are independent Weibull distributed 
$$
X_1,\ldots,X_n \vert \theta, k \overset{\mathrm{iid}}{\sim} \mathrm{Weibull}(\theta,k).
$$
The exponential distribution used in Exercise 2.2 is the special case with $k=1$ where the failure (death) rate is constant over time. For $k<1$ the failure rate is descreasing with time and $k>$ gives an increasing failure rate over time. <br>
Use the prior $\theta \sim \mathrm{Gamma}(\alpha,\beta)$ independent of $k \sim \mathrm{Expon}(\gamma)$. Set $\alpha=3$ and $\beta=300$ as in Exercise 2.2 and $\gamma = 1$ so that $\mathbb{E}(k) = 1$, i.e. the prior is centering the Weibull model on the Exponential model.

::: {#prob:weibull_lung .callout-note icon="false" collapse="true"}
## Solution Exercise 2.3
From Exercise 2.1, we know that the posterior distribution is 
$$\theta \sim \mathrm{Gamma}(\alpha + n_u, \beta + \sum\nolimits_{u \in \mathcal{U}} x_u),$$
where $n_u$ is the number of uncensored observations and $\mathcal{U}$ is the set of observation indices for the uncensored data.

The following code plots the prior, likelihood (normalized) and posterior over a grid of values for $\theta$. Note that the data is so much stronger than the prior that the posterior is virtually identical to the likelihood, which is why the normalized likelihood is not visible in the plot.

```{r}
#| output: false
library(tidyverse) # loads data manipulation and visualization packages
library(survival) # loads the lung cancer data as `lung`
colors = c("#6C8EBF", "#c0a34d", "#780000","#007878","#B5C6DF","#EADAAA","#AE6666")
```

```{r}
# Summarize the data needed for the posterior, filter out censored data
data_summary <- lung %>% filter(status == 2) %>% summarize(n = n(), sum_x = sum(time))
```

```{r}
# Set up prior hyperparameters
alpha_prior <- 3   # shape parameter
beta_prior <- 300  # rate parameter

# Compute posterior hyperparameters
alpha_post <- alpha_prior + data_summary$n  
beta_post <- beta_prior + data_summary$sum_x   
```

```{r}
# Plot the prior and posterior densities, and the (normalized) likelihood as a bon 
thetaGrid <- seq(0, 0.03, length.out = 1000)
prior_density <- dgamma(thetaGrid, shape = alpha_prior, rate = beta_prior)
likelihood_density <- dgamma(thetaGrid, shape = data_summary$n, rate = data_summary$sum_x)
posterior_density <- dgamma(thetaGrid, shape = alpha_post, rate = beta_post)

df <- data.frame(
  thetaGrid = thetaGrid, 
  prior = prior_density, 
  likelihood = likelihood_density,
  posterior = posterior_density
)

df_long <- df %>% pivot_longer(-thetaGrid, names_to = "density_type", values_to = "density")

# Plot using ggplot2
ggplot(df_long) +
  aes(x = thetaGrid, y = density, color = density_type) +
  geom_line() +
  scale_colour_manual(
    breaks = c("prior", "likelihood", "posterior"), 
    values = c(colors[2], colors[1], colors[3])) +
  labs(title = "Exercise 2.2", x = expression(theta), y = "Density", color = "") + 
  theme_minimal()
```

:::


::: {#prob:expon_lung_b .callout-note icon="false" collapse="true"}
## Solution Exercise 2.2 b.

The likelihood for all data, censored and uncensored, is 
$$
\begin{align}
p(x_1,\ldots,x_n \vert \theta) & = \prod_{i=1}^n p(x_i \vert \theta) \\
& = \prod_{u \in \mathcal{U}} p(x_u \vert \theta) \prod_{c \in \mathcal{C}} p(x_c \vert \theta) \\
& = \prod_{u \in \mathcal{U}} p(x_u \vert \theta) \prod_{c \in \mathcal{C}} \left(1 - F(x_c \vert \theta)\right) 
\end{align}
$$
where $\mathcal{U}$ and $\mathcal{C}$ are the sets of observation indicies for the uncensored and censored data, respectively. The likelihood for the uncensored data (the first product) is the same as before
$$
\prod_{u \in \mathcal{U}} p(x_u \vert \theta) = \prod_{u \in \mathcal{U}} \theta e^{-\theta x_u} = \theta^{n_u} e^{-\theta\sum_{u \in \mathcal{U}} x_u},
$$
where $n_u$ is the number of uncensored observations. The likelihood contribution for each observation in the censored set (the second product) is the survival function
$$
\mathrm{Pr}(X \geq x_c) = 1 - F(x_c \vert \theta),
$$
where $F(x_c \vert \theta) = 1 - e^{-x_c \theta}$ is the cumulative distribution function of the exponential distribution evaluated at $x_c$.

So the likelihood function is
$$
p(x_1,\ldots,x_n \vert \theta) = \theta^n e^{-\theta\sum_{u \in \mathcal{U}} x_u} \times e^{-\theta \sum_{u \in \mathcal{U}} x_c} = \theta^{n_u} e^{-\theta\sum_{i = 1}^n x_i}.
$$
where one should note that $\theta$ is raised to the number of uncensored observations, $n_u$ while the sum in the exponential term includes both uncensored and censored observations.

By Bayes' theorem, the posterior distribution is again a Gamma distribution
$$
\begin{align}
p(\theta \vert x_1,\ldots,x_n) & \propto p(x_1,\ldots,x_n \vert \theta)p(\theta) \\
& \propto \theta^{n_u} e^{-\theta\sum_{i = 1}^n x_i} \times \theta^{\alpha-1}e^{-\beta\theta} \\
& = \theta^{\alpha + n_u - 1} e^{ -\theta(\beta + \sum_{i = 1}^n x_i)},
\end{align}
$$
which we recognize as proportional to the following Gamma distribution
$$
\theta \vert x_1,\ldots,x_n \sim \mathrm{Gamma}(\alpha + n_u,\beta + \sum\nolimits_{i=1}^n x_i).
$$ 

The code below plots both:

- the posterior from the previous exercise (a) with only the uncensored data and 
- the posterior from with all data.

The posterior with all data is more informative and concentrates on smaller $\theta$ values. Since smaller $\theta$ values correspond to longer expected survival times, this is makes sense since the censored patients were still alive at the end of the study.

```{r}
# Summarize the data needed for the posterior, grouped by `status`:
data_summary <- lung %>% group_by(status) %>% summarize(n = n(), sum_x = sum(time))
```

```{r}
# Set up prior hyperparameters
alpha_prior <- 3   # shape parameter
beta_prior <- 300  # rate parameter

# Compute posterior hyperparameters - only uncensored data
alpha_post_u <- alpha_prior + data_summary$n[2] # second row is uncensored data (status = 2)  
beta_post_u <- beta_prior + data_summary$sum_x[2] # sum over uncensored observations

# Compute posterior hyperparameters - all data
alpha_post_all <- alpha_prior + data_summary$n[2] # note: this is still n_u 
beta_post_all <- beta_prior + sum(data_summary$sum_x) # sum over all observations   
```

```{r}
# Plot the prior and the two posterior densities 
thetaGrid <- seq(0, 0.03, length.out = 1000)
prior_density <- dgamma(thetaGrid, shape = alpha_prior, rate = beta_prior)
posterior_density_u <- dgamma(thetaGrid, shape = alpha_post_u, rate = beta_post_u)
posterior_density_all <- dgamma(thetaGrid, shape = alpha_post_all, rate = beta_post_all)


df <- data.frame(
  thetaGrid = thetaGrid, 
  prior = prior_density, 
  posterior_uncensored = posterior_density_u,
  posterior_all = posterior_density_all
)

df_long <- df %>% pivot_longer(-thetaGrid, names_to = "density_type", values_to = "density")

# Plot using ggplot2
ggplot(df_long) +
  aes(x = thetaGrid, y = density, color = density_type) +
  geom_line() +
  scale_colour_manual(
    breaks = c("prior", "posterior_uncensored", "posterior_all"), 
    values = c(colors[2], colors[3], colors[4])) +
  labs(title = "Exercise 2.2", x = expression(theta), y = "Density", color = "") + 
  theme_minimal()
```

:::
