---
title: "Modeling galaxy velocities with a Dirichlet process mixture model"
author: "Mattias Villani"
draft: true
format:
  html:
    fig-align: center
    toc: true
bibliography: ../notebook.bib
csl: ../apa.csl
---

## Introduction

The **galaxy velocity data**[^1] consists of the velocities of 82 galaxies (measured in 1000km/s, after transformation). The scientific question is whether or not the velocities of the galaxies cluster around a few values, which would indicate that the galaxies are part of a few superclusters[^2].

[^1]: Postman, M., Huchra, J. P. and Geller, M. J. (1986) Probes of large-scale structures in the Corona Borealis region. *Astronomical Journal* **92**, 1238–1247.

[^2]: [Roeder, K. (1990). *Density estimation with confidence sets exemplified by superclusters and voids in the galaxies*. Journal of the American Statistical Association, 85(411), 617-624](https://doi.org/10.1080/01621459.1990.10474918)

The analysis here first explores the **Dirichlet process** as a prior directly on the distribution (cdf) of the velocities before moving on to the more interesting **Dirichlet process mixture (DPM) model**. The DP mixture makes it possible to learn about the number of mixture components in the data, i.e. to explore the possibility of superclusters of galaxies.

```{r}
#| code-fold: true
#| code-summary: "Code for loading packages and plot settings."
#| output: false
#| warning: false
library(MASS)
library(ggplot2)
library(patchwork)

colors <- c("#6C8EBF", "#c0a34d", "#780000", "#007878",
            "#b5c6df", "#eadaaa", "#AE6666", "#4CA0A0", "#bf9d6c")

theme_set(theme_bw(base_size = 11) + theme(panel.grid = element_blank()))
set.seed(123)
```

The data are available in the `galaxies` dataset in the `MASS` package in R. We inspect the data by plotting a histogram and a kernel density estimate using cross-validation to select the optimal bandwidth. At least three clusters are visible in the histogram, while the kernel density estimate suggest up to seven clusters.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false
data(galaxies, package = "MASS")
velocity <- galaxies / 1000
xlimits  <- c(5, 40)
xGrid    <- seq(xlimits[1], xlimits[2], length.out = 1000)
n        <- length(velocity)

kde_fit <- density(velocity, bw = "ucv")

p_data <- ggplot(data.frame(x = velocity), aes(x = x)) +
  geom_histogram(aes(y = after_stat(density)), fill = colors[6],
                 color = colors[2], linewidth = 0.5, bins = 20) +
  geom_line(data = data.frame(x = kde_fit$x, y = kde_fit$y),
            aes(x = x, y = y), color = colors[1], linewidth = 1.5) +
  xlim(xlimits) +
  labs(x = "Velocity (km/s)", y = "density", title = "Kernel density estimate")
p_data
```

## Dirichlet process prior for the cdf

We first analyze the cumulative distribution function (cdf) of the velocities using a Dirichlet process prior.

#### Empirical cdf

The empirical cumulative distribution function (cdf) for the data can be computed by

```{r}
empirical_cdf <- function(x, xlimits) {
  xs <- sort(x)
  ps <- (1:length(xs)) / length(xs)
  xs <- c(xlimits[1], xs, xlimits[2])
  ps <- c(0, ps, 1)
  list(xs = xs, ps = ps)
}

ecdf_vals <- empirical_cdf(velocity, xlimits)
p_ecdf <- ggplot(data.frame(x = ecdf_vals$xs, y = ecdf_vals$ps), aes(x, y)) +
  geom_step(color = "black", direction = "hv") +
  labs(x = "Velocity (km/s)", y = "CDF", title = "Empirical cdf")
p_ecdf
```

When we later combine the empirical cdf with a Dirichlet process prior to form the posterior distribution it will be convenient to have a proper distribution object for the empirical cdf. We therefore approximate it with a mixture of normals with components with tiny ($10^{-15}$) variances, effectively making it a mixture of Dirac spikes at the data points with equal weights $1/n$. The plot below shows that the approximation is virtually identical to the true empirical cdf.

```{r}
# CDF of the normal-mixture approximation to Fn (equivalent to the empirical CDF)
cdf_Fn <- function(x_vals) {
  sapply(x_vals, function(xv) mean(pnorm(xv, mean = velocity, sd = 1e-10)))
}

p_ecdf +
  geom_line(data = data.frame(x = xGrid, y = cdf_Fn(xGrid)),
            aes(x, y), color = colors[2], inherit.aes = FALSE,
            linewidth = 1)
```

#### Prior distribution

Let $P$ denote the unknown distribution for the galaxy velocities. We use a Dirichlet process for $P$, denoted by $P \sim \mathrm{DP}(\alpha_0P_0)$. The $P_0$ is the so called **base measure** and is our best prior guess (prior expected value) of $P$. We take $P_0$ to be the $N(30,3^2)$ distribution. The scalar parameter $\alpha_0$ is the prior concentration or **prior precision**, which is inversely related to the variance around $P_0$. We start with low prior concentration $\alpha_0=1$ and later explore the effect of changing $\alpha_0$.

```{r}
#| output: false
mu0_dp    <- 20
sigma0_dp <- 3   # P0 = Normal(20, 3)
alpha0    <- 1
```

At this stage we would like to simulate draws from the DP prior and plot the resulting distribution realizations, to get a feeling for the effect of $\alpha_0$. The so called stick-breaking representation of the Dirichlet process is a constructive characterization that makes it easy to simulate from a DP. The **stick-breaking representation** of the Dirichlet process is an infinite mixture of point masses $\delta_{\theta_k}$ at locations $\theta_k$ with weights $\omega_k$:

$$
\begin{align}
  P(\cdot) &= \sum_{k=1}^{\infty}\omega_{k}\delta_{\theta_{k}} \nonumber \\
  \omega_{k} &= V_{k}\prod_{\ell<k}(1-V_{\ell})  \nonumber \\
  V_{k} &\overset{\mathrm{iid}}{\sim}\mathrm{Beta}(1,\alpha_0) \nonumber \\
  \theta_{k} &\overset{\mathrm{iid}}{\sim}P_{0}
\end{align}
$$

The function below simulates from the Dirichlet process using the stick-breaking construction. The input `r_P0` is a function that draws from the base measure $P_0$, and `epsilon` is the remaining stick length when the simulation terminates.

```{r}
#| output: false
sim_dir_process <- function(r_P0, alpha0, epsilon = 0.001) {
  stick_left <- 1
  theta <- numeric(0)
  omega <- numeric(0)
  while (stick_left > epsilon) {
    theta_k    <- r_P0()
    V_k        <- rbeta(1, 1, alpha0)
    omega_k    <- V_k * stick_left
    stick_left <- stick_left * (1 - V_k)
    theta      <- c(theta, theta_k)
    omega      <- c(omega, omega_k)
  }
  sort_idx <- order(theta)
  list(theta = theta[sort_idx], omega = omega[sort_idx])
}

# Efficient CDF evaluation when theta is sorted
dp_cdf <- function(dp, x_vals) {
  cum_omega <- c(0, cumsum(dp$omega))
  bin_idx   <- findInterval(x_vals, dp$theta)
  cum_omega[bin_idx + 1]
}
```

Here are a few draws from the prior for three different choices of $\alpha_0$.

```{r}
n_draws <- 3
r_P0 <- function() rnorm(1, mu0_dp, sigma0_dp)

plts_prior <- lapply(c(1, 10, 100), function(a0) {
  draws_df <- do.call(rbind, lapply(1:n_draws, function(i) {
    dp <- sim_dir_process(r_P0, a0)
    data.frame(x = xGrid, y = dp_cdf(dp, xGrid), draw = factor(i))
  }))
  ggplot(draws_df, aes(x, y, color = draw)) +
    geom_line() +
    scale_color_manual(values = colors[1:n_draws]) +
    labs(title = bquote(paste("Prior draws for ", alpha[0], " = ", .(a0))),
         x = "x", y = "CDF") +
    theme(legend.position = "none")
})
plts_prior[[1]] + plts_prior[[2]] + plts_prior[[3]] + plot_layout(ncol = 3)
```

We can simulate a large number of draws and construct $95\%$ prior pointwise credible intervals for $P$, showing the variation in the prior. Here are the (pointwise) prior credible intervals for $\alpha_0=1$.

```{r}
n_sim  <- 10000
alpha0 <- 1
prior_draws <- matrix(0, length(xGrid), n_sim)
for (i in 1:n_sim) {
  dp <- sim_dir_process(r_P0, alpha0)
  prior_draws[, i] <- dp_cdf(dp, xGrid)
}
prior_q_cdf <- apply(prior_draws, 1, quantile, probs = c(0.025, 0.975))

df_prior_ci <- data.frame(
  x  = xGrid,
  P0 = pnorm(xGrid, mu0_dp, sigma0_dp),
  lo = prior_q_cdf[1, ],
  hi = prior_q_cdf[2, ]
)
ggplot(df_prior_ci, aes(x = x)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = colors[2], alpha = 0.2) +
  geom_line(aes(y = P0), color = colors[2]) +
  labs(title = bquote(alpha[0] == .(alpha0)), x = "x", y = "CDF")
```

#### Posterior distribution

The posterior distribution for $P$ is also a Dirichlet process

$$
P \vert x_1,\ldots, x_n \sim \mathrm{DP}(\alpha_nP_n)
$$where the posterior precision is $\alpha_n = \alpha_0 + n$ and

$$
P_n = \frac{\alpha_0}{\alpha_n}P_0 + \frac{n}{\alpha_n}F_n
$$

is a mixture of the prior mean and the empirical cdf. We compute the posterior by combining the two components with weights $\alpha_0/\alpha_n$ and $n/\alpha_n$:

```{r}
alpha_n <- alpha0 + n
cdf_Pn <- function(x_vals, a0, n_obs) {
  an <- a0 + n_obs
  (a0 / an) * pnorm(x_vals, mu0_dp, sigma0_dp) +
    (n_obs / an) * ecdf(velocity)(x_vals)
}
```

We can now plot the prior-to-posterior mapping. Here is the code and plot for $\alpha_0=100$.

```{r}
n_sim  <- 5000
alpha0 <- 100

prior_draws <- matrix(0, length(xGrid), n_sim)
for (i in 1:n_sim) {
  dp <- sim_dir_process(r_P0, alpha0)
  prior_draws[, i] <- dp_cdf(dp, xGrid)
}
prior_q_cdf <- apply(prior_draws, 1, quantile, probs = c(0.025, 0.975))

alpha_n_val <- alpha0 + n
r_Pn <- function() {
  if (runif(1) < alpha0 / alpha_n_val) rnorm(1, mu0_dp, sigma0_dp)
  else sample(velocity, 1)
}

post_draws <- matrix(0, length(xGrid), n_sim)
for (i in 1:n_sim) {
  dp <- sim_dir_process(r_Pn, alpha_n_val)
  post_draws[, i] <- dp_cdf(dp, xGrid)
}
post_q_cdf <- apply(post_draws, 1, quantile, probs = c(0.025, 0.975))

df_dp <- data.frame(
  x        = xGrid,
  Fn       = ecdf(velocity)(xGrid),
  P0       = pnorm(xGrid, mu0_dp, sigma0_dp),
  Pn       = cdf_Pn(xGrid, alpha0, n),
  prior_lo = prior_q_cdf[1, ],
  prior_hi = prior_q_cdf[2, ],
  post_lo  = post_q_cdf[1, ],
  post_hi  = post_q_cdf[2, ]
)

ggplot(df_dp, aes(x = x)) +
  geom_ribbon(aes(ymin = prior_lo, ymax = prior_hi), fill = colors[2], alpha = 0.2) +
  geom_ribbon(aes(ymin = post_lo,  ymax = post_hi),  fill = colors[3], alpha = 0.2) +
  geom_line(aes(y = Fn), color = colors[1]) +
  geom_line(aes(y = P0), color = colors[2]) +
  geom_line(aes(y = Pn), color = colors[3]) +
  labs(title = bquote(alpha[0] == .(alpha0)), x = "x", y = "CDF")
```

## Dirichlet process mixture

The posterior distribution for $P$ in the above analysis is almost surely discrete. This is an undesirable consequence of the Dirichlet process prior whenever $P$ is thought to be a continuous distribution.

However, the almost sure discreteness of the DP can be turned into a useful feature if the DP is used as mixing measure in a mixture model. A mixture model for the random variable $X$ is of the form

$$
p(x) = \int p(x \vert \mathbf{\theta})dP(\mathbf{\theta})
$$

where $p(x\vert\theta)$ is the kernel distribution with parameter vector $\mathbf{\theta}$ and $P(\mathbf{\theta})$ is the mixing distribution. The notation $dP(\mathbf{\theta})$ comes from measure theory and $P(\cdot)$ is a probability measure. This more general version of the integral makes it possible for $P(\cdot)$ be both a discrete and continuous probability distribution, or a mixture of the two types. There are three basic types of mixture distributions:

1.  **Continuous mixtures** where $P(\mathbf{\theta})$ is a continuous distribution with density $p(\theta)$ so we can write the more familiar integral $p(x) = \int p(x \vert \mathbf{\theta})p(\mathbf{\theta})d\mathbf{\theta}$. Example: the student $t$ density is a mixture of Gaussian kernels where the variance parameter of the kernel is mixed with respect to a scaled-inverse-chi-square distribution.

2.  **Finite discrete mixtures**\
    $$p(x) = \sum_{k=1}^K \omega_k \cdot N(x\vert \mu_k, \sigma_k^2),$$\
    where the kernel is again Gaussian, but the mixing distribution is a mixture of $K$ point masses for the mean and variance of the kernel. Since the mixing distribution $P(\cdot)$ is discrete here, the integral reduces to a sum.

3.  **Infinite discrete mixtures**\
    $$p(x) = \sum_{k=1}^\infty \omega_k \cdot N(x\vert \mu_k, \sigma_k^2),$$\
    where the most popular version is the Dirichlet Process Mixture (DPM) model, which is obtained when $P(\mathbf{\theta})$ is a *random distribution* following a Dirichlet process. The Dirichlet Process Mixture (DPM) model can be written\
    $$
    \begin{aligned}
    x_i \vert \mathbf{\theta}_i &\overset{iid}{\sim} p(x_i \vert \mathbf{\theta}_i) \\  \mathbf{\theta}_i \vert P &\overset{iid}{\sim} P \nonumber \\
    P &\sim \mathrm{DP}(\alpha_0 P_0)
    \end{aligned}
    $$

Since realizations from the DP is almost surely discrete there will usually be several observations sharing the same parameter value $\mathbf{\theta}_j^{\star}$. In the case with a Gaussian kernel, the parameter vector is $\mathbf{\theta} =(\mu,\sigma^2)$ and the base measure $P_0$ is taken to be the conjugate prior:

$$
\begin{aligned}
\mu \vert \sigma^2 \sim N\bigg(\mu_0,\frac{\sigma^2}{\kappa_0}\bigg) \\
\sigma^2 \sim \mathrm{Inv}-\chi^2(\nu_0, \sigma_0^2)
\end{aligned}
$$In the Gibbs sampler presented below, we will need to compute the posterior of $\mathbf{\theta} =(\mu,\sigma^2)$ for all the observations currently allocated to a mixture component. This little helper function computes the posterior parameters $\mu_n,\kappa_n,\sigma_n^2$ and $\nu_n$ from a given data vector `x`:

```{r}
#| output: false
post_normal_model <- function(x, mu0, sigma2_0, kappa0, nu0) {
  if (length(x) == 0) return(list(mu = mu0, sigma2 = sigma2_0, kappa = kappa0, nu = nu0))
  xbar    <- mean(x)
  n       <- length(x)
  kappa_n <- kappa0 + n
  w       <- n / kappa_n
  mu_n    <- w * xbar + (1 - w) * mu0
  nu_n    <- nu0 + n
  sumsqr  <- sum((x - xbar)^2)
  sigma2_n <- (nu0 * sigma2_0 + sumsqr + (kappa0 * n / kappa_n) * (xbar - mu0)^2) / nu_n
  list(mu = mu_n, sigma2 = sigma2_n, kappa = kappa_n, nu = nu_n)
}
```

The Scaled inverse Chi-square distribution used as the prior on $\sigma^2$ can be simulated using the chi-squared distribution: if $X \sim \chi^2(\nu)$ then $\nu\tau^2/X \sim \mathrm{Inv}\text{-}\chi^2(\nu, \tau^2)$.

```{r}
#| output: false
r_scaled_inv_chi2 <- function(nu, tau2) nu * tau2 / rchisq(1, nu)
```

The complete blocked Gibbs sampler for the DPM model with Gaussian components is given below. The sampler was developed in [@ishwaran2001gibbs]; see also [@villani_bayesbook]. The prior precision $\alpha_0$ can be fixed, or assigned a $\alpha_0 \sim \mathrm{Gamma}(g,h)$ prior and learned from the data. If `alpha0 = NULL` then $\alpha_0$ is estimated.

```{r}
#| output: false
gibbs_dpm_normals <- function(x, K, n_iter, n_burn, P0,
                              alpha0 = NULL, g = 1, h = 1) {
  n      <- length(x)
  mu0    <- P0[1]; sigma2_0 <- P0[2]; kappa0 <- P0[3]; nu0 <- P0[4]

  mu     <- if (K == 1) mean(x) else seq(min(x), max(x), length.out = K)
  sigma2 <- rep(var(x) / K, K)
  omega  <- rep(1 / K, K)
  V      <- rep(0, K)
  z      <- rep(1L, n)

  estimate_alpha0 <- is.null(alpha0)
  if (estimate_alpha0) {
    alpha0       <- 1
    alpha0_draws <- numeric(n_iter)
  } else {
    alpha0_draws <- NULL
  }

  mu_draws     <- matrix(0, n_iter, K)
  sigma2_draws <- matrix(0, n_iter, K)
  omega_draws  <- matrix(0, n_iter, K)
  K_occupied   <- integer(n_iter)

  for (j in 1:(n_burn + n_iter)) {

    # Update component allocations
    log_lik <- sweep(
      sapply(1:K, function(k) dnorm(x, mu[k], sqrt(sigma2[k]), log = TRUE)),
      2, log(omega), "+"
    )
    log_lik <- log_lik - apply(log_lik, 1, max)
    probs   <- exp(log_lik)
    probs   <- probs / rowSums(probs)
    z       <- apply(probs, 1, function(p) sample.int(K, 1, prob = p))

    n_in_comps <- tabulate(z, nbins = K)

    # Update component parameters
    for (k in 1:K) {
      post      <- post_normal_model(x[z == k], mu0, sigma2_0, kappa0, nu0)
      sigma2[k] <- r_scaled_inv_chi2(post$nu, post$sigma2)
      mu[k]     <- rnorm(1, post$mu, sqrt(sigma2[k] / post$kappa))
    }

    # Update component weights via stick-breaking
    stick_length <- 1
    for (k in 1:(K - 1)) {
      a_post <- 1 + n_in_comps[k]
      b_post <- alpha0 + sum(n_in_comps[(k + 1):K])
      V[k]   <- rbeta(1, a_post, b_post)
      omega[k]     <- V[k] * stick_length
      stick_length <- stick_length * (1 - V[k])
    }
    omega[K] <- stick_length

    # Update alpha0 (Gamma prior with scale parameterization)
    if (estimate_alpha0) {
      alpha0 <- rgamma(1, shape = g + K + 1,
                       rate  = h - sum(log(1 - V[1:(K - 1)])))
    }

    if (j > n_burn) {
      idx              <- j - n_burn
      mu_draws[idx, ]     <- mu
      sigma2_draws[idx, ] <- sigma2
      omega_draws[idx, ]  <- omega
      K_occupied[idx]     <- sum(n_in_comps > 0)
      if (estimate_alpha0) alpha0_draws[idx] <- alpha0
    }
  }

  list(mu_draws = mu_draws, sigma2_draws = sigma2_draws,
       omega_draws = omega_draws, K_occupied = K_occupied,
       alpha0_draws = alpha0_draws)
}
```

After setting the prior hyperparameters and upper bound of the number of mixtures components, we run the Gibbs sampler for `n_burn=1000` iterations as burn-in and then another `n_iter = 10000` for posterior inference.

```{r}
#| output: false
P0     <- c(mu0 = 20, sigma2_0 = 1, kappa0 = 0.1, nu0 = 3)
alpha0 <- 1
K      <- 30
n_burn <- 1000
n_iter <- 10000
fit    <- gibbs_dpm_normals(velocity, K, n_iter, n_burn, P0, alpha0)
```

A helper function computes the posterior mean density by averaging mixture pdfs over a thinned set of MCMC draws.

```{r}
#| output: false
compute_post_mean_pdf <- function(fit, xGrid, K, plt_iter) {
  pdf_mat <- sapply(plt_iter, function(i) {
    rowSums(sapply(1:K, function(k)
      fit$omega_draws[i, k] *
        dnorm(xGrid, fit$mu_draws[i, k], sqrt(fit$sigma2_draws[i, k]))))
  })
  rowMeans(pdf_mat)
}

plot_K_bar <- function(K_occupied) {
  K_max  <- max(K_occupied)
  df_bar <- data.frame(
    k    = 0:K_max,
    freq = sapply(0:K_max, function(k) mean(K_occupied == k))
  )
  ggplot(df_bar, aes(x = k, y = freq)) +
    geom_bar(stat = "identity", fill = colors[5],
             color = colors[1], linewidth = 0.5) +
    labs(x = "Number of components", y = "density",
         title = "Posterior for number of components")
}
```

Plot the posterior mean of the fitted density for $\alpha_0=1$.

```{r}
plt_iter     <- seq(1, n_iter, by = 10)
post_mean_pdf <- compute_post_mean_pdf(fit, xGrid, K, plt_iter)

p_data +
  geom_line(data = data.frame(x = xGrid, y = post_mean_pdf),
            aes(x = x, y = y), color = colors[3], linewidth = 1.5,
            inherit.aes = FALSE) +
  labs(title = bquote(paste("Post. mean density, ", alpha[0], " = ", .(alpha0))))
```

The posterior for the number of occupied mixture components has most of its mass on 5-10 components, with a mode at seven components:

```{r}
plot_K_bar(fit$K_occupied)
```

For $\alpha_0 = 10$ we get

```{r}
#| output: false
alpha0  <- 10
fit10   <- gibbs_dpm_normals(velocity, K, n_iter, n_burn, P0, alpha0)
```

```{r}
pdf10 <- compute_post_mean_pdf(fit10, xGrid, K, plt_iter)
p_post10 <- p_data +
  geom_line(data = data.frame(x = xGrid, y = pdf10),
            aes(x = x, y = y), color = colors[3], linewidth = 1.5,
            inherit.aes = FALSE) +
  labs(title = bquote(paste("Post. mean density, ", alpha[0], " = ", .(alpha0))))
p_post10 + plot_K_bar(fit10$K_occupied)
```

For $\alpha_0=100$

```{r}
#| output: false
alpha0  <- 100
fit100  <- gibbs_dpm_normals(velocity, K, n_iter, n_burn, P0, alpha0)
```

```{r}
pdf100 <- compute_post_mean_pdf(fit100, xGrid, K, plt_iter)
p_post100 <- p_data +
  geom_line(data = data.frame(x = xGrid, y = pdf100),
            aes(x = x, y = y), color = colors[3], linewidth = 1.5,
            inherit.aes = FALSE) +
  labs(title = bquote(paste("Post. mean density, ", alpha[0], " = ", .(alpha0))))
p_post100 + plot_K_bar(fit100$K_occupied)
```

and finally when $\alpha_0$ is estimated using a Gamma(1,1) prior:

```{r}
#| output: false
fit_est <- gibbs_dpm_normals(velocity, K, n_iter, n_burn, P0)
```

```{r}
pdf_est <- compute_post_mean_pdf(fit_est, xGrid, K, plt_iter)
p_post_est <- p_data +
  geom_line(data = data.frame(x = xGrid, y = pdf_est),
            aes(x = x, y = y), color = colors[3], linewidth = 1.5,
            inherit.aes = FALSE) +
  labs(title = expression(paste("Post. mean density, ", alpha[0], " ~ Gamma(1,1)")))
p_post_est + plot_K_bar(fit_est$K_occupied)
```

The posterior for $\alpha_0$ is

```{r}
ggplot(data.frame(alpha0 = fit_est$alpha0_draws), aes(x = alpha0)) +
  geom_histogram(aes(y = after_stat(density)), fill = colors[7],
                 color = colors[3], alpha = 0.7, linewidth = 0.5, bins = 50) +
  labs(x = expression(alpha[0]), y = "density",
       title = expression(paste("Posterior for ", alpha[0])))
```
